AFQ-Insight

### 2019-12-10 reports/manuscript/sections/methods.tex
Add error term to definition of linear model and use y-hat, beta-hat afterward to represent predictions

After scaling and imputation, the tractometry data and target phenotypical data
can be organized in a linear model:

\begin{equation}
y = \mathbf{X} \cdot \beta,
\label{eq:lm}
\end{equation}

where $y$ is the phenotype -- categorical, such as a clinical diagnosis, or
continuous numerical, such as the subject's age. The tractometry data is
represented in the feature matrix $\mathbf{X}$, with rows corresponding to
different subjects, and columns corresponding to diffusion measures at different
nodes within each tract. The relationship between tractometric features and the
phenotypic target is characterized by the coefficients in $\beta$. In general,
the feature matrix $\mathbf{X}$ has dimensions $S \times (T \cdot N \cdot M)$,
where $S$ is the number of subjects, $T$ is the number of white matter tracts,
$N$ is the number of nodes in each tract, and $M$ is the number of diffusion
metrics calculated at each node. Typically, $T = 20$, $N = 100$, and $2 \le M
\le 8$, resulting in $\sim 4,000 - 16,000$  features. Conversely, many dMRI
studies have between a few dozen and a few hundred subjects, yielding a feature
matrix that is wide and short. Even in cases where more than a thousand subjects
are measured (e.g., in the Human Connectome Project, where 1,200 subjects were
measured \cite{VanEssen2012}), the problem is ill-posed: the high dimensionality
of this data requires regularization to avoid overfitting and generate
interpretable results.

Here, we propose that in addition to regularizing the coefficients in $\beta$,
we can also capitalize on our knowledge of the group structure of the tract
profile features in $\mathbf{X}$. The tract-metric combinations form a natural
grouping. For example, we expect that MD features within the left arcuate
fasciculus will co-vary across individuals. Likewise for FA values within the
right corticospinal tract (CST) and so on. This group structure is represented
in Fig~\ref{fig:group-structure}, which depicts the linear model $y = \mathbf{X}
\cdot \beta$. Thus, we seek a regularization approach that will fit a linear
model with anatomically-grouped covariates, where we expect to observe both
groupwise sparsity, where the number of groups (tract/metric combinations) with
at least one non-zero coefficients is small, as well as within-group sparsity,
where the number of non-zero coefficients within each non-zero group is small.
The sparse group lasso (SGL)
is a penalized regression technique that satisfies exactly these
criteria\cite{simon2013sparse}. It solves for a coefficient vector
$\hat{\beta}$ that satisfies
\begin{equation}
    \hat{\beta} = \min_\beta \frac{1}{2}
    ||y - \displaystyle \sum_{\ell = 1}^{G}
    \mathbf{X}^{(\ell)} \cdot \beta^{(\ell)}||_2^2
    + \lambda_1 \displaystyle \sum_{\ell = 1}^{G}
    \sqrt{p_\ell} ||\beta^{(\ell)}||_2
    + \lambda_2 ||\beta||_1,
    \label{eq:sgl}
\end{equation}

where $G$ is the number of groups $\mathbf{X}^{(\ell)}$ is the submatrix
of $\mathbf{X}$ corresponding to group $\ell$, $\beta^{(\ell)}$ is
the coefficient vector for group $\ell$ and $p_\ell$ is the length of
$\beta^{(\ell)}$. In the tractomtetry setting, $G = T \cdot M$ and
$\forall \ell: p_\ell = 100$. The first term is the mean square error
loss, $L_{\text{mse}}$, as in the standard linear regression framework.
The second and third terms encourage groupwise sparsity and overall
sparsity, respectively. If $\lambda_1 = 0$ and $\lambda_2 = 1$, the
SGL reduces to the traditional lasso\cite{tibshirani1996regression}.
Conversely, if $\lambda_1 = 1$ and $\lambda_2 = 0$, the SGL reduces to
the group lasso\cite{yuan2006model}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{dMRI_group_structure.png}
    \caption{{\bf dMRI group structure.}
        The phenotypical target data and tractometric features can be organized
    into a linear model, $y = \mathbf{X} \cdot \beta$. The feature matrix
    $\mathbf{X}$ is color-coded to reveal a natural group structure: the left
    (orange) group contains $k$ features from the inferior fronto-occipital
    fascicle (IFOF), the middle (green) group contains $k$ features from the
    corpus callosum, and the right (blue) group contains $k$ features from the
    uncinate. The coefficients in $\beta$ follow the same natural grouping.
    Fascicle image reproduced with permission from Ref~\cite{yeatman2012tract}
    Figure 1.}
    \label{fig:group-structure}
\end{figure}

@@

Often, the target variable $y$ is not in the domain in which the linear model
can be best fit to it. Equation \ref{eq:lm} can be slightly modified as:
\begin{equation}
y = f^-1(\mathbf{X} \cdot \beta),
\label{eq:lm}
\end{equation}
where the transformation function $f^{-1}$ characterizes the transform applied
to the data before fitting the linear coefficients. For example, an often-used
transform is the logarithmic transform:
\begin{equation}
f(y) = \log_n(y)
\label{eq:log_nonlinearity}
\end{equation}
In this case, the model is parametrized by one additional fit parameter, $n$.

\subsubsection*{Classification of categorical $y$}
When the phenotypical target variable is categorical, as in the case of
explaining or predicting the presence of a clinical diagnosis, the SGL is
readily adapted to logistic regression, where the probability of a target
variable belonging to an arbitrary defined ``true'' class is the logistic
function of the result of the linear model,
\begin{equation}
    p(y = 1) = \frac{1}{1 + \exp(\mathbf{X}\cdot \beta)},
    \label{eq:logit}
\end{equation}
or equivalently, the mean squared error loss function in Eq~\eqref{eq:sgl} is
replaced with the cross-entropy loss, which for binary classification is the
negative log likelihood of the SGL classifier giving the ``true'' label:
\begin{equation}
    L_{\text{mse}} \rightarrow L_{\log} =
    -\left(y \log(p) + (1 - y) \log(1 - p)\right).
    \label{eq:logloss}
\end{equation}

@@

To objectively evaluate the model and guard against over-fitting, we used a
nested cross-validation scheme, depicted in Fig~\ref{fig:cross-val} for the
categorical classification case. The input data is shuffled and then decomposed
into $k$ batches, hereafter referred to as folds. For the ALS dataset we used
$k=10$ and for the WH dataset $k=5$. For each unique fold, we hold that fold out
as the test\textsubscript{outer} set and let the remaining data comprise the
train\textsubscript{outer} set, with the subscript indicating the depth of the
nested decomposition. We further decompose each train\textsubscript{outer} set
into three folds, and again for each unique fold, we hold out that fold as the
test\textsubscript{inner} set and let the remaining data comprise the
train\textsubscript{inner} set. At level-1 of the decomposition, we fit an SGL
model using fixed regularization meta-parameters $\lambda_1$ and $\lambda_2$,
training the model using train\textsubscript{inner} and evaluating the fit on
test\textsubscript{inner}. We find the optimal values for $\lambda_1$ and
$\lambda_2$ using hyperoptimizatin. This is implemented using the hyperopt
library's \verb|fmin| function\cite{Bergstra_2015} with a tree of Parzen
estimators search algorithm\cite{bergstra2011algorithms}. For continuous
numerical $y$, \verb|fmin| searches for meta-parameter values that minimize the
median absolute error. This can also be done minimizing the root of the mean
squared error (RMSE) or to maximizing the coefficient of determination ($R^2$).
For binary categorical $y$ \verb|fmin| seeks to maximize the classification
accuracy. This can also be done maximizing the area under the receiver operating
curve (ROC AUC) or the average precision. Using hyperoptimization, we find
optimal regularization parameters and $\hat{\beta}$ for each
train\textsubscript{outer} set and then use those to predict values for data in
test\textsubscript{outer}. Thus each subject in the dataset has a predicted
phenotype derived from a model that never saw its particular subject's data.

>>>>

After scaling and imputation, the tractometry data and target
phenotypical data can be organized in a linear model:
\begin{equation}
    y = \mathbf{X} \beta + \epsilon,
    \label{eq:lm}
\end{equation}
where $y$ is the phenotype -- categorical, such as a clinical diagnosis,
or continuous numerical, such as the subject's age. The tractometry
data is represented in the feature matrix $\mathbf{X}$, with rows
corresponding to different subjects, and columns corresponding
to diffusion measures at different nodes within each tract. The
relationship between tractometric features and the phenotypic target is
characterized by the coefficients in $\beta$. The error term, $\epsilon$
is an unobserved random variable that captures the error in the model.
We denote our prediction of the targer phenotype as $\hat{y}$ and the
coefficients that produce this prediction as $\hat{\beta}$, so that
\begin{equation}
    \hat{y} = \mathbf{X} \hat{\beta},
    \label{eq:lm-approx}
\end{equation}
without the error term, $\epsilon$. In general, the feature matrix
$\mathbf{X}$ has dimensions $S \times (T \times N \times M)$, where $S$
is the number of subjects, $T$ is the number of white matter tracts,
$N$ is the number of nodes in each tract, and $M$ is the number of
diffusion metrics calculated at each node. Typically, $T = 20$, $N =
100$, and $2 \le M \le 8$, resulting in $\sim 4,000 - 16,000$ features.
Conversely, many dMRI studies have between a few dozen and a few
hundred subjects, yielding a feature matrix that is wide and short.
Even in cases where more than a thousand subjects are measured (e.g.,
in the Human Connectome Project, where 1,200 subjects were measured
\cite{VanEssen2012}), the problem is ill-posed: the high dimensionality
of this data requires regularization to avoid overfitting and generate
interpretable results.

Here, we propose that in addition to regularizing the coefficients
in $\hat{\beta}$, we can also capitalize on our knowledge of the
group structure of the tract profile features in $\mathbf{X}$. The
tract-metric combinations form a natural grouping. For example, we
expect that MD features within the left arcuate fasciculus will
co-vary across individuals. Likewise for FA values within the right
corticospinal tract (CST) and so on. This group structure is represented
in Fig~\ref{fig:group-structure}, which depicts the linear model
$\hat{y} = \mathbf{X} \hat{\beta}$. Thus, we seek a regularization
approach that will fit a linear model with anatomically-grouped
covariates, where we expect to observe both groupwise sparsity, where
the number of groups (tract/metric combinations) with at least one
non-zero coefficients is small, as well as within-group sparsity, where
the number of non-zero coefficients within each non-zero group is small.
The sparse group lasso (SGL) is a penalized regression technique that
satisfies exactly these criteria\cite{simon2013sparse}. It solves for a
coefficient vector
$\hat{\beta}$ that satisfies
\begin{equation}
    \hat{\beta} = \min_\beta \frac{1}{2}
    ||y - \displaystyle \sum_{\ell = 1}^{G}
    \mathbf{X}^{(\ell)} \beta^{(\ell)}||_2^2
    + \lambda_1 \displaystyle \sum_{\ell = 1}^{G}
    \sqrt{p_\ell} ||\beta^{(\ell)}||_2
    + \lambda_2 ||\beta||_1,
    \label{eq:sgl}
\end{equation}
where $G$ is the number of groups $\mathbf{X}^{(\ell)}$ is the submatrix
of $\mathbf{X}$ corresponding to group $\ell$, $\beta^{(\ell)}$ is
the coefficient vector for group $\ell$ and $p_\ell$ is the length of
$\beta^{(\ell)}$. In the tractomtetry setting, $G = T \times M$ and
$\forall \ell: p_\ell = 100$. The first term is the mean square error
loss, $L_{\text{mse}}$, as in the standard linear regression framework.
The second and third terms encourage groupwise sparsity and overall
sparsity, respectively. If $\lambda_1 = 0$ and $\lambda_2 = 1$, the
SGL reduces to the traditional lasso\cite{tibshirani1996regression}.
Conversely, if $\lambda_1 = 1$ and $\lambda_2 = 0$, the SGL reduces to
the group lasso\cite{yuan2006model}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{dMRI_group_structure.png}
    \caption{{\bf dMRI group structure.}
        The phenotypical target data and tractometric features can
        be organized into a linear model, $\hat{y} = \mathbf{X}
        \hat{\beta}$. The feature matrix $\mathbf{X}$ is color-coded
        to reveal a natural group structure: the left (orange) group
        contains $k$ features from the inferior fronto-occipital
        fascicle (IFOF), the middle (green) group contains $k$ features
        from the corpus callosum, and the right (blue) group
        contains $k$ features from the uncinate. The coefficients in
        $\hat{\beta}$ follow the same natural grouping. Fascicle image
        reproduced with permission from Ref~\cite{yeatman2012tract}
        Figure 1.
    }
    \label{fig:group-structure}
\end{figure}

@@

Often, the target variable $y$ is not in the domain in which the linear
model can be best fit to it. Equation \ref{eq:lm-approx} can be slightly
modified as:
\begin{equation}
    \hat{y} = f^-1(\mathbf{X} \hat{\beta}),
    \label{eq:lm-transform}
\end{equation}
where the transformation function $f^{-1}$ characterizes the transform
applied to the data before fitting the linear coefficients. For example,
an often-used transform is the logarithmic transform:
\begin{equation}
    f(\hat{y}) = \log_n(\hat{y})
    \label{eq:log-nonlinearity}
\end{equation}
In this case, the model is parametrized by one additional fit parameter,
$n$.

\subsubsection*{Classification of categorical $y$}
When the phenotypical target variable is categorical, as in the case of
explaining or predicting the presence of a clinical diagnosis, the SGL is
readily adapted to logistic regression, where the probability of a target
variable belonging to an arbitrary defined ``true'' class is the logistic
function of the result of the linear model,
\begin{equation}
    p(\hat{y} = 1) = \frac{1}{1 + \exp(\mathbf{X} \hat{\beta})},
    \label{eq:logit}
\end{equation}
or equivalently, the mean squared error loss function in Eq~\eqref{eq:sgl} is
replaced with the cross-entropy loss, which for binary classification is the
negative log likelihood of the SGL classifier giving the ``true'' label:
\begin{equation}
    L_{\text{mse}} \rightarrow L_{\log} =
    -\left(y \log(p) + (1 - y) \log(1 - p)\right).
    \label{eq:logloss}
\end{equation}

@@

To objectively evaluate the model and guard against over-fitting, we used a
nested cross-validation scheme, depicted in Fig~\ref{fig:cross-val} for the
categorical classification case. The input data is shuffled and then decomposed
into $k$ batches, hereafter referred to as folds. For the ALS dataset we used
$k=10$ and for the WH dataset $k=5$. For each unique fold, we hold that fold out
as the test\textsubscript{outer} set and let the remaining data comprise the
train\textsubscript{outer} set, with the subscript indicating the depth of the
nested decomposition. We further decompose each train\textsubscript{outer} set
into three folds, and again for each unique fold, we hold out that fold as the
test\textsubscript{inner} set and let the remaining data comprise the
train\textsubscript{inner} set. At level-1 of the decomposition, we fit an SGL
model using fixed regularization meta-parameters $\lambda_1$ and $\lambda_2$,
training the model using train\textsubscript{inner} and evaluating the fit on
test\textsubscript{inner}. We find the optimal values for $\lambda_1$ and
$\lambda_2$ using hyperoptimization, implemented using the hyperopt
library's \verb|fmin| function\cite{Bergstra_2015} with a tree of Parzen
estimators search algorithm\cite{bergstra2011algorithms}. For continuous
numerical $y$, \verb|fmin| searches for meta-parameter values that minimize the
median absolute error. This can also be done minimizing the root of the mean
squared error (RMSE) or to maximizing the coefficient of determination ($R^2$).
For binary categorical $y$ \verb|fmin| seeks to maximize the classification
accuracy. This can also be done maximizing the area under the receiver operating
curve (ROC AUC) or the average precision. Using hyperoptimization, we find
optimal regularization parameters and $\hat{\beta}$ for each
train\textsubscript{outer} set and then use those to predict values for data in
test\textsubscript{outer}. Thus each subject in the dataset has a predicted
phenotype derived from a model that never saw its particular subject's data.
